A non-additive generalization of probability theory \\for quantum mechanics and beyond

### 2025-07-03 Articles/2024-NonAdditiveProbability/NonAdditiveProbability.tex
NonAddititiveProbability: more work on the standard probability section

\section{Additive measure and Choquet theory}

[This section has to make clear in what extra conditions on ensembles spaces Choquet theory can be applied.]

% Intro paragraph that says quickly how to connect to Choquet theory and its limitations. It should be a quick summary of the whole section
Since ensemble spaces are a convex subset of a locally convex topological vector space, they can be connected to Choquet theory.

% Quick review of Choquet theory
Let us first give a summary of the main results in Choquet theory that are interesting for our purposes. First of all, let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $\mu : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_E f d\mu$ for every continuous linear functional $f$ on $V$. The Choquet theorem states:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}

>>>>

\section{Probability measures and Choquet theory}

\textbf{Classical probability measures.} In the classical case, ensembles are defined as probability measures of the space of all possible states $X$. In the discrete case, like for roll of a die, $X$ is a set endowed with the discrete topology\footnote{Physically, the discrete topology corresponds to the ability to experimentally verify or falsify each case. That is, we are able to recognize whether a particular die landed or didn't land on a particular face.} and, consequently, the power set as the $\sigma$-algebra. In the continuous case, $X$ will be classical phase space endowed with the usual topology of a manifold\footnote{Physically, the topology of a manifold corresponds to the ability to label each possible case through a set of continuous variable which are verifiable up to finite but arbitrarily small precision. That is, we are able to tell whether a state is within a particular range of position and momentum.} and the standard Borel sets as the $\sigma$-algebra. In either case, a probability measure is a map $p : \Sigma_X \to [0,1]$ that assign to each event of the $\sigma$-algebra (i.e. to an empirical statements) the probability to find it to be true. The probability measure is additive, meaning that if two events $A, B \in \Sigma_X$ are mutually exclusive (i.e. $A \cap B = \emptyset$) then the probability of the union is the sum of the probability of the respective events:
\begin{equation}
	p(A\cup B) = p(A) + p(B).
\end{equation}

\textbf{Absolute continuity of probability measures.} While each classical ensemble is a probability distribution, not all probability distribution can be thought as a valid ensemble. For example, we could have a probability measure to be wholly defined over a single point of phase space, which would correspond to the ability to prepare a system in the same exact place with the same exact momentum. Not only this is not feasible experimentally, which violates the whole premise of having objects that map to physically meaningful entities, but it would have to have entropy equal minus infinity. To address this, we add a requirement that is both mathematically natural and physically meaningful. Let $\mu : \Sigma_X \to [0, + \infty]$ the measure on $X$ that quantifies the count of states in each region. In the discrete case this is simply the counting measure (i.e. three possible faces of the die corresponds to three possible states), while in the continuous case this corresponds to the Liouville measure (i.e. volumes in phase space correspond to the count of states). This measure is additive over mutually exclusive events. The requirement is that we can assign non-zero probability only to events that have a non-zero count of states. That is, if for a particular event $A$ we have $p(A) \neq 0$, then we must have that $\mu(A) \neq 0$ as well. Mathematically, $p$ is absolutely continuous with respect to $\mu$, and this is exactly the case in which a probability density $\rho = \frac{dp}{d\mu}$ can be defined.\footnote{In the general case, we can still assign non-zero probability to a point of a manifold, provided that the topology and the measure $\mu$ consistently allow for it. Physically, it means that those points are ``special'' both in terms of experimental verifiability and entropy/count of states. While the general framework allows for it, it is not crucial for our discussion here.} This is another example in which a physical requirement (probability must be zero if there are no possible states), once properly spelled out (absolute continuity), justifies common physical intuition (we assume there is a probability density).

% Intro paragraph that says quickly how to connect to Choquet theory and its limitations. It should be a quick summary of the whole section

\textbf{Choquet theory.} Having reviewed the use of probability measures in classical mechanics, we can ask the more general question: when can we represent ensembles in a generic ensemble space as a probability measure? This question is treated by Choquet theory, though in not the full generality that we need it. Let us therefore review the main results of Choquet theory, before discussing how it connects to ensembles spaces.

Mathematically, the goal of Choquet theory is to represent a convex set as a probability measure over the extreme points. Let $E$ be a convex subset of a locally convex space $V$. We say that $\ens \in E$ is represented by a probability measure $\mu : \Sigma_{E} \to [0,1]$ if $f(\ens) = \int_E f d\mu$ for every continuous linear functional $f$ on $V$. The Choquet theorem states:
\begin{thrm}[Choquet]
	Let $E$ be a metrizable compact convex subset of a locally convex space $V$, and let $\ens \in E$. Then there exists a probability measure $\mu_{\ens}$ on $X$ which represents $\ens$ and is supported by the extreme points of $X$.
\end{thrm}
Since we saw that an ensemble space is convex subset of a locally convex topological vector space, the relevance of Choquet theory is straight forward.
